---
title: "CS5801 Coursework Supporting analysis Template Proforma"
author: "STUDENT_ID"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
version: 1
---

# 0. Instructions 
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(car)
library(MASS)




```

# 1. Organise and clean the data

## 1.1 Subset the data into the specific dataset allocated
 
```{r}
# Student ID and offset

SID <- STUDENT_ID
SIDoffset <- (SID %% 50) + 1
SIDoffset

# Load dataset

load("carbon-footprint.Rda")
ls()

# Subset allocated observations

Author_STUDENT_ID_mydf_Support <- df[seq(from = SIDoffset,
to = nrow(df),
by = 50), ]

# Initial checks

dim(Author_STUDENT_ID_mydf_Support)
head(Author_STUDENT_ID_mydf_Support)
tail(Author_STUDENT_ID_mydf_Support)
str(Author_STUDENT_ID_mydf_Support)
summary(Author_STUDENT_ID_mydf_Support)

# Save allocated subset

write.csv(Author_STUDENT_ID_mydf_Support,
"Author_STUDENT_ID_mydf_Support.csv",
row.names = FALSE)

```


## 1.2 Data quality plan

``` {r}
str(Author_STUDENT_ID_mydf_Support)
summary(Author_STUDENT_ID_mydf_Support)
colSums(is.na(Author_STUDENT_ID_mydf_Support))

if ("home_type" %in% names(Author_STUDENT_ID_mydf_Support))
table(Author_STUDENT_ID_mydf_Support$home_type)

if ("diet_type" %in% names(Author_STUDENT_ID_mydf_Support))
table(Author_STUDENT_ID_mydf_Support$diet_type)

if ("uses_public_transport" %in% names(Author_STUDENT_ID_mydf_Support))
table(Author_STUDENT_ID_mydf_Support$uses_public_transport)

if ("recycles_regularly" %in% names(Author_STUDENT_ID_mydf_Support))
table(Author_STUDENT_ID_mydf_Support$recycles_regularly)




```



## 1.3 Data quality analysis findings


```{r}
# Missing income

if ("income" %in% names(Author_STUDENT_ID_mydf_Support)) {
sum(is.na(Author_STUDENT_ID_mydf_Support$income))
summary(Author_STUDENT_ID_mydf_Support$income)
}

# Duplicate recycling variables

if (all(c("recycles_regularly","recycles_regularly_question")
%in% names(Author_STUDENT_ID_mydf_Support))) {
table(Author_STUDENT_ID_mydf_Support$recycles_regularly)
table(Author_STUDENT_ID_mydf_Support$recycles_regularly_question)
with(Author_STUDENT_ID_mydf_Support,
table(recycles_regularly, recycles_regularly_question))
}

# Range check for annual_energy_consumption

if ("annual_energy_consumption" %in% names(Author_STUDENT_ID_mydf_Support)) {
summary(Author_STUDENT_ID_mydf_Support$annual_energy_consumption)
Author_STUDENT_ID_mydf_Support[
Author_STUDENT_ID_mydf_Support$annual_energy_consumption < 0, ]
}





```
 
## 1.4 Data cleaning  
 

```{r}
# Create cleaned dataset used in submission

mydf <- Author_STUDENT_ID_mydf_Support

# Issue 1: Mean imputation for income (used in submission)

if ("income" %in% names(mydf)) {
income_mean <- mean(mydf$income, na.rm = TRUE)
mydf$income[is.na(mydf$income)] <- income_mean
summary(mydf$income)
}

# Alternative: Median imputation

mydf_income_median <- Author_STUDENT_ID_mydf_Support
if ("income" %in% names(mydf_income_median)) {
income_median <- median(mydf_income_median$income, na.rm = TRUE)
mydf_income_median$income[is.na(mydf_income_median$income)] <- income_median
summary(mydf_income_median$income)
}

# Issue 2: Drop duplicate recycling column (used in submission)

if ("recycles_regularly_question" %in% names(mydf)) {
mydf$recycles_regularly_question <- NULL
}
names(mydf)

# Alternative: Re-code Yes/No recycling variable

mydf_recycle_recode <- Author_STUDENT_ID_mydf_Support
if ("recycles_regularly_question" %in% names(mydf_recycle_recode)) {
mydf_recycle_recode$recycle_from_text <-
mydf_recycle_recode$recycles_regularly_question
mydf_recycle_recode$recycle_from_text[
mydf_recycle_recode$recycle_from_text == "Yes"] <- 1
mydf_recycle_recode$recycle_from_text[
mydf_recycle_recode$recycle_from_text == "No"] <- 0
mydf_recycle_recode$recycle_from_text <-
as.numeric(mydf_recycle_recode$recycle_from_text)

with(mydf_recycle_recode,
table(recycles_regularly, recycle_from_text))
}

# Issue 3: Negative annual_energy_consumption -> NA then mean imputation

if ("annual_energy_consumption" %in% names(mydf)) {
mydf$annual_energy_consumption[
mydf$annual_energy_consumption < 0] <- NA

energy_mean <- mean(mydf$annual_energy_consumption, na.rm = TRUE)
mydf$annual_energy_consumption[
is.na(mydf$annual_energy_consumption)] <- energy_mean

summary(mydf$annual_energy_consumption)
}
# Alternative: Remove observations with invalid annual_energy_consumption values

mydf_energy_removed <- Author_STUDENT_ID_mydf_Support

mydf_energy_removed <- mydf_energy_removed[
  mydf_energy_removed$annual_energy_consumption >= 0 |
    is.na(mydf_energy_removed$annual_energy_consumption),
]

# Check dimensions after removal
dim(mydf_energy_removed)

# Check remaining range
summary(mydf_energy_removed$annual_energy_consumption)


# Save final cleaned dataset

write.csv(mydf, "Author_STUDENT_ID_mydf.csv", row.names = FALSE)


```


# 2. Exploratory Data Analysis (EDA)


  
```{r}
str(mydf)
summary(mydf)

table(mydf$home_type)
table(mydf$diet_type)
table(mydf$uses_public_transport)
table(mydf$recycles_regularly)

prop.table(table(mydf$recycles_regularly))

hist(mydf$carbon_footprint)
hist(mydf$annual_energy_consumption)
boxplot(mydf$income)

boxplot(carbon_footprint ~ home_type, data = mydf)
boxplot(carbon_footprint ~ diet_type, data = mydf)

plot(mydf$annual_energy_consumption, mydf$carbon_footprint)



```


# 3. Modelling



```{r}
start_model <- lm(carbon_footprint ~ home_type +
annual_energy_consumption +
household_size +
income,
data = mydf)
summary(start_model)

model_cf_1 <- lm(carbon_footprint ~ home_type +
annual_energy_consumption +
household_size,
data = mydf)
summary(model_cf_1)

model_cf_2 <- lm(carbon_footprint ~ home_type +
annual_energy_consumption,
data = mydf)
summary(model_cf_2)

final_model <- lm(carbon_footprint ~ home_type +
annual_energy_consumption,
data = mydf)
summary(final_model)

plot(final_model)

coef(final_model)
head(final_model$fitted.values)
head(final_model$residuals)

anova(start_model, model_cf_2)




```


# 4. Modelling another dependent variable


```{r}
recycle_model <- glm(recycles_regularly ~ home_type +
annual_energy_consumption,
data = mydf,
family = binomial)
summary(recycle_model)

plot(recycle_model)

coef(recycle_model)
summary(recycle_model$fitted.values)

exp(coef(recycle_model))



```

# References  
Agresti, A. (2019) An introduction to categorical data analysis. 3rd edn. Hoboken, NJ: Wiley.

Brunel University London (2025) CS5801: Quantitative Data Analysis – Lecture notes and laboratory materials. Brunel University London.

Brunel University London (2025) CS5702: Modern Data – Lecture notes and laboratory materials. Brunel University London.

James, G., Witten, D., Hastie, T. and Tibshirani, R. (2021) An introduction to statistical learning: With applications in R. 2nd edn. New York: Springer.

MarinStatsLectures (n.d.) MarinStatsLectures – R Programming & Statistics YouTube channel. YouTube. Available at: https://www.youtube.com/c/MarinStatsLectures-RProgrammingStatistics
 

R Core Team (2024) R: A language and environment for statistical computing. Vienna: R Foundation for Statistical Computing.


R Documentation (n.d.) The base package. Available at: https://stat.ethz.ch/R-manual/R-devel/library/base/html/
 

Wickham, H. (2016) ggplot2: Elegant graphics for data analysis. New York: Springer.

